Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-09-23 introduction.tex
remarks reviewer 1 fixed or in margin (blue)

Various extensions of language models for handling infinite alphabets have been studied.
%words carrying data values in an infinite domain (e.g. integers) e.g. data processing
Some automata with memory extensions
allow restricted storage and comparison of input symbols
%and correspond with logics,
(see~\cite{Segoufin06csl} for a survey)
\florent{register: skip refs and details, add Mikolaj recent}
using pebbles for marking positions~\cite{NevenSchwentickVianu04FSMinfinite},
registers~\cite{KaminskiFrancez94},
or %computations in 2 steps
the possibility to compute on subsequences
with the same attribute values~\cite{Bojanczyk11FO2}. % data words automata.
%
%for the and verification of infinite state systems
%(model checkers: alphabet = long bit-vectors)
%...For the representation of  in model checking, verification and
%
The automata at the core of model checkers
compute on input symbols represented by large bitvectors~\cite{Vardi07ciaa} %\cite{BaierKatoen08MC}
(sets of assignments of Boolean variables) %propositional variables)
and, in practice,  %implementation
each transition accepts a set of such symbols (instead of an individual symbol),
represented by Boolean Formulas or Binary Decision Diagrams.
%
Following a similar idea, % and generalizing,
in symbolic automata (\SA)~\cite{dAntoniVeanes17CAV,dAntoni21CACM},
transitions are guarded by predicates over infinite  domains.
With appropriate closure conditions on the sets of such predicates, % (alphabet theories),
all the good properties enjoyed over finite alphabets are preserved.

Other extensions of language models  %(automata and grammars)
help in dealing with non-determinism by computing weight values.
With an ambiguous grammar, there may exist several derivations
(\emph{abstract syntax trees} -- AST) % \ie the result of an analyze.
yielding one input word. % structuring a word,
The association of one weight value %in semiring domain
to each AST permits to select a best one (or $n$ bests). % a fixed number of bests
% = ...a ranking of ASTs
%This is roughly the principle of \emph{weighted parsing} approaches
\florent{WARNING: \cite{MorbitzVogler19weighted-parsing} much is more general that weighted parsing}
In \emph{weighted language models}~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing},
like \eg probabilistic context-free grammars % (CFG),
and weighted automata (\WA) \cite{Droste09handbook},
a weight is associated to each transition rule, % production rule
and the rule's weights can be combined with an
associative product operator~$\otimes$ to yield the weight of an AST.
A second operator~$\oplus$
is moreover used to resolve the ambiguity raised by the existence
of several (in general exponentially many) AST
associated to a given input word.
Typically, $\oplus$  selects the best of two weight values.
%Intuitively,~$\oplus$ selects, or ranking, the syntax trees.
The weight domain, equipped with these two operators is typically 
a \emph{semiring} %$\Semiring$
where $\oplus$ can be extended to infinite sums,
%\cite{Eilenberg74automata}
such as the Viterbi semiring and the tropical min-plus algebra% % see Figure~\ref{fig:semirings}.
%of domain $\mathbb{R}^+ \cup \{ +\infty\}$,
%where $\oplus$ is min and $\otimes$ is plus .
%by ranking
%making the weight domain a semiring.
%Some efficient specialized parsing algorithms~\cite{Huang05kbest} have been proposed in this context
%% models represented as hypergraphs \cite{Huang05kbest}
%in order to compute the $n$ best syntax trees of a given input word without having to enumerate them all.
%Generally based on dynamic programming, these algorithms rely on
%additional algebraic properties of~$\Semiring$.
%-- see \eg~\cite{Huang05kbest} for some NLP applications.
%The extraction of $n$ best list is useful
%the problem: quantitative parsing or symbolic parsing
%of parsing of words over infinite input alphabet.

In this paper, we present a framework for weighted parsing over infinite input alphabets.
It is based on \emph{symbolic weighted} finite states language models (\swM),
generalizing the Boolean guards of~\SA %in transitions
to functions into an arbitrary semiring,
and generalizing also~\WA, by handling infinite alphabets (Figure~\ref{fig:hierarchy}).
%
In short, a transition rule $q \xrightarrow{\phi} q'$, from state $q$ to $q'$,
%\florent{c'est un résumé de la Fig. \ref{fig:hierarchy}}
%input symbols are variables, %(parameters)
is labeled by a function~$\phi$ associating to every input symbol $a$ a weight value $\phi(a)$
in a semiring $\Semiring$.

@@

The framework relies on several language models:
symbolic-weighted automata (\SWA),
transducers (\SWT), and pushdown automata
with a visibility restriction~\cite{AlurMadhusudan09nested} (\SWVPA).
A \SWT defines a distance $T(s, t)$ between finite words $s$ and $t$
over infinite alphabets. % following~\cite{Mohri03ijfcs}.
%\wrt an edit distance
A  \SWVPA operates sequentially on \emph{nested words}~\cite{AlurMadhusudan09nested},
structured with markup symbols (parentheses) and describing linearizations of trees.
A \SWVPA $A$ associates a weight value $A(t)$ % \in \Semiring$
to a given nested word $t$, which is itsef the linearization of a weighted AST. %representing a parse tree.
%
Then, given an input word $s$, the \emph{SW-parsing} problem aims at
finding $t$ minimizing
$T(s, t) \otimes A(t)$, called the distance between $s$ and $A$ in~\cite{Mohri03ijfcs},
\wrt the ranking defined by $\oplus$. Like weighted-parsing
methods~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing},
\florent{WARNING: \cite{MorbitzVogler19weighted-parsing} much is more general that weighted parsing}
our approach proceeds in two steps.
The first step is an intersection
(Bar-Hillel construction~\cite{GruneJacobs08parsing})
\florent{chap. intersection in~\cite{GruneJacobs08parsing}}
where, given a \SWT $T$, a \SWVPA $A$, and an input word $s$,
a \SWVPA $B$ %$A_{T, s}$
is built, such that for all~$t$,
$B(t) = T(s, t) \otimes A(t)$. % $A_{T, s}(t) = T(s, t) \otimes A(t)$.
In the second step, a best AST $t$ is found by applying to $B$ %$A_{T, s}$
a best search algorithm similar to the shortest distance
in graphs~\cite{Mohri02semiring,Huang05kbest}.
%In expressiveness, they are equivalent to weighted CFG.
%and can be used in a general approach for parsing over infinite input alphabets.
%
%Let $A$ be a \SWVPA, associating $A(w) \in \Semiring$
%to a given a nested word $w$ (representing a parse tree),
%and let a \SWT compute a distance $d$, in $\Semiring$,
%between 2 strings over respectively an infinite input alphabet and the
%same (infinite) alphabet of $A$.
%Then, the problem of Symbolic Weighted Parsing is,
%given an input string $s$, to find a nested word $w$ minimizing
%(according to the ranking defined by $\oplus$)
%the distance $d(s, w) \otimes A(w)$ between $s$ and $A$,
%as defined in~\cite{Mohri03ijfcs}.

% First one general edit-distance is defined by a weighted word
% transducer~\cite{Mohri}
% %Symbolic automata transducers are extended models~\cite{VeanesdAntoniJACM}
% %dealing with infinite set of input symbols...
% value in a semiring...
\florent{expressiveness: VPA have restricted equality test.
        comparable to pebble automata? $\to$ conclusion}
The main contributions of the paper are:
($i$)~the introduction of automata, \SWA, transducers, \SWT (Section~\ref{section:SWA}),
and visibly pushdown automata \SWVPA (Section~\ref{sec:SWVPA}),
generalizing the corresponding classes of symbolic and weighted models,
%for the weighted computation on (nested) words over infinite alphabets;
($ii$)~a polynomial best-search algorithm for \SWVPA, %(Section~\ref{sec:best})
% a framework for parsing over infinite alphabets,
and ($iii$)~a uniform framework (Section~\ref{sec:parsing}) for parsing over infinite alphabets,
the keys to which are
($iii.a$)~the \SWT-based definition of generic edit distances between input and output (yield) words,
and ($iii.b$)~the use, for convenience in this context,
of nested words, and \SWVPA,
instead of syntax trees and grammars. %(\S~\ref{sec:trees}).

>>>>

Various extensions of language models for handling infinite alphabets have been studied.
%words carrying data values in an infinite domain (e.g. integers) e.g. data processing
Some automata 
\reviews{1) and grammar with storage?}
with memory extensions
allow restricted storage and comparison of input symbols
%and correspond with logics,
(see~\cite{Segoufin06csl} for a survey)
\florent{register: skip refs and details, add Mikolaj recent}
using pebbles for marking positions~\cite{NevenSchwentickVianu04FSMinfinite},
registers~\cite{KaminskiFrancez94},
or %computations in 2 steps
the possibility to compute on subsequences
with the same attribute values~\cite{Bojanczyk11FO2}. % data words automata.
%
%for the and verification of infinite state systems
%(model checkers: alphabet = long bit-vectors)
%...For the representation of  in model checking, verification and
%
The automata at the core of model checkers
compute on input symbols represented by large bitvectors~\cite{Vardi07ciaa} %\cite{BaierKatoen08MC}
(sets of assignments of Boolean variables) %propositional variables)
and, in practice,  %implementation
each transition accepts a set of such symbols (instead of an individual symbol),
represented by Boolean Formulas or Binary Decision Diagrams.
%
Following a similar idea, % and generalizing,
in symbolic automata (\SA)~\cite{dAntoniVeanes17CAV,dAntoni21CACM},
transitions are guarded by predicates over infinite  domains.
With appropriate closure conditions on the sets of such predicates, % (alphabet theories),
all the good properties enjoyed over finite alphabets are preserved.

Other extensions of language models  %(automata and grammars)
help in dealing with non-determinism by computing weight values.
With an ambiguous grammar, 
\reviews{1) jump from automata to grammars is uneasy and not really necessary}
there may exist several derivations
(\emph{abstract syntax trees} -- AST) % \ie the result of an analyze.
yielding one input word. % structuring a word,
The association of one weight value %in semiring domain
to each AST permits to select a best one (or $n$ bests). % a fixed number of bests
% = ...a ranking of ASTs
%This is roughly the principle of \emph{weighted parsing} approaches
\florent{introduce pb "weighted parsing" and ref.\cite{MorbitzVogler19weighted-parsing}.} 
In \emph{weighted language models}~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing},
like \eg probabilistic context-free grammars % (CFG),
and weighted automata (\WA) \cite{Droste09handbook},
a weight is associated to each transition rule, % production rule
and the rule's weights can be combined with an
associative product operator~$\otimes$ to yield the weight of an AST.
\reviews{1) AST for grammar, run for WA}
A second operator~$\oplus$
is moreover used to resolve the ambiguity raised by the existence
of several (in general exponentially many) AST
associated to a given input word.
Typically, $\oplus$  selects the best of two weight values.
%Intuitively,~$\oplus$ selects, or ranking, the syntax trees.
The weight domain, equipped with these two operators is typically 
a \emph{semiring} %$\Semiring$
where $\oplus$ can be extended to infinite sums,
%\cite{Eilenberg74automata}
such as the Viterbi semiring and the tropical min-plus algebra% % see Figure~\ref{fig:semirings}.
%of domain $\mathbb{R}^+ \cup \{ +\infty\}$,
%where $\oplus$ is min and $\otimes$ is plus .
%by ranking
%making the weight domain a semiring.
%Some efficient specialized parsing algorithms~\cite{Huang05kbest} have been proposed in this context
%% models represented as hypergraphs \cite{Huang05kbest}
%in order to compute the $n$ best syntax trees of a given input word without having to enumerate them all.
%Generally based on dynamic programming, these algorithms rely on
%additional algebraic properties of~$\Semiring$.
%-- see \eg~\cite{Huang05kbest} for some NLP applications.
%The extraction of $n$ best list is useful
%the problem: quantitative parsing or symbolic parsing
%of parsing of words over infinite input alphabet.

In this paper, we present a framework for weighted parsing over infinite input alphabets.
It is based on \emph{symbolic weighted} finite states language models (\swM),
generalizing the Boolean guards of~\SA %in transitions
to functions into an arbitrary semiring,
and generalizing also~\WA, by handling infinite alphabets (Figure~\ref{fig:hierarchy}).
\reviews{cite \cite{Herrmann16dlt,Herrmann20phd}}
%
In short, a transition rule $q \xrightarrow{\phi} q'$, from state $q$ to $q'$,
%\florent{c'est un résumé de la Fig. \ref{fig:hierarchy}}
%input symbols are variables, %(parameters)
is labeled by a function~$\phi$ associating to every input symbol $a$ a weight value $\phi(a)$
in a semiring $\Semiring$.

@@

The framework relies on several language models:
symbolic-weighted automata (\SWA),
transducers (\SWT), and pushdown automata
with a visibility restriction~\cite{AlurMadhusudan09nested} (\SWVPA).
A \SWT defines a distance $T(s, t)$ between finite words $s$ and $t$
over infinite alphabets. % following~\cite{Mohri03ijfcs}.
%\wrt an edit distance
A  \SWVPA operates sequentially on \emph{nested words}~\cite{AlurMadhusudan09nested},
structured with markup symbols (parentheses) and describing linearizations of trees.
A \SWVPA $A$ associates a weight value $A(t)$ % \in \Semiring$
to a given nested word $t$, which is itsef the linearization of a weighted AST. %representing a parse tree.
%
Then, given an input word $s$, the \emph{SW-parsing} problem aims at
finding $t$ minimizing
$T(s, t) \otimes A(t)$, called the distance between $s$ and $A$ in~\cite{Mohri03ijfcs},
\wrt the ranking defined by $\oplus$. Like weighted-parsing
methods~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing},
\reviews{1) WARNING: \cite{MorbitzVogler19weighted-parsing} incomparable to our def. weighted parsing}
our approach proceeds in two steps.
The first step is an intersection
(Bar-Hillel construction~\cite{GruneJacobs08parsing})
\reviews{1) also \cite{NederhofSatta03ParsingIntersection} pasing as intersection}
\florent{chap. intersection in~\cite{GruneJacobs08parsing}}
where, given a \SWT $T$, a \SWVPA $A$, and an input word $s$,
a \SWVPA $B$ %$A_{T, s}$
is built, such that for all~$t$,
$B(t) = T(s, t) \otimes A(t)$. % $A_{T, s}(t) = T(s, t) \otimes A(t)$.
In the second step, a best AST $t$ is found by applying to $B$ %$A_{T, s}$
a best search algorithm similar to the shortest distance
in graphs~\cite{Mohri02semiring,Huang05kbest}.
%In expressiveness, they are equivalent to weighted CFG.
%and can be used in a general approach for parsing over infinite input alphabets.
%
%Let $A$ be a \SWVPA, associating $A(w) \in \Semiring$
%to a given a nested word $w$ (representing a parse tree),
%and let a \SWT compute a distance $d$, in $\Semiring$,
%between 2 strings over respectively an infinite input alphabet and the
%same (infinite) alphabet of $A$.
%Then, the problem of Symbolic Weighted Parsing is,
%given an input string $s$, to find a nested word $w$ minimizing
%(according to the ranking defined by $\oplus$)
%the distance $d(s, w) \otimes A(w)$ between $s$ and $A$,
%as defined in~\cite{Mohri03ijfcs}.

% First one general edit-distance is defined by a weighted word
% transducer~\cite{Mohri}
% %Symbolic automata transducers are extended models~\cite{VeanesdAntoniJACM}
% %dealing with infinite set of input symbols...
% value in a semiring...
\florent{expressiveness: VPA have restricted equality test.
        comparable to pebble automata? $\to$ conclusion}
The main contributions of the paper are:
\reviews{1) not $i$! see response.txt}
($i$)~the introduction of automata, \SWA, transducers, \SWT (Section~\ref{section:SWA}),
and visibly pushdown automata \SWVPA (Section~\ref{sec:SWVPA}),
generalizing the corresponding classes of symbolic and weighted models,
%for the weighted computation on (nested) words over infinite alphabets;
($ii$)~a polynomial best-search algorithm for \SWVPA, %(Section~\ref{sec:best})
% a framework for parsing over infinite alphabets,
and ($iii$)~a uniform framework (Section~\ref{sec:parsing}) for parsing over infinite alphabets,
the keys to which are
($iii.a$)~the \SWT-based definition of generic edit distances between input and output (yield) words,
and ($iii.b$)~the use, for convenience in this context,
of nested words, and \SWVPA,
instead of syntax trees and grammars. %(\S~\ref{sec:trees}).

### 2021-09-23 main.tex
minor modifs
remarks reviewer 1 fixed or in margin (blue)

% comments in margin with \todo command
\usepackage[colorinlistoftodos]{todonotes} % option disable to hide todos
% command todo : options
% - noline: no line connecting the note with the place in the text
%   where the note occurs in the latex code
% - fancyline : curved arrow from note to inssertion point
% - inline: place a todonote inside the text instead of in the margin
% - author=name
\newcommand{\florent}[1]{\todo[noline,size=\tiny,color=yellow!40]{#1}}
%\renewcommand{\florent}[1]{} % uncomment for final
\newcommand{\lydia}[1]{\todo[noline,size=\tiny,color=red!30]{#1}}
%\renewcommand{\lydia}[1]{} % uncomment for final
\newcommand{\philippe}[1]{\todo[noline,size=\tiny,color=green!30]{#1}}
%\renewcommand{\philippe}[1]{} % uncomment for final

@@

%\begin{example}
%We can define with a \SWT the computation of a similarity measure
%between timed sequences
%similar to dynamic time warping (DTW).
%
%Let $\Semiring$  be the tropical (\emph{min-plus}) semiring of Figure~\ref{fig:semirings} and
%let $\Sigma = \Delta = \mathbb{R}_+$ be sets of timestamps.
%We consider a \SWT with one state $q$ and transitions
%$\wei_{11}(q, d, d', q) =
% \wei_{01}(q, d, d', q) =
% \wei_{10}(q, d, d', q) = |d' - d|$,
%for all $d, d' \in \mathbb{R}_+$.
%% needs reading input/output symbols even by epsilon-transitions.
%The recursive definition of $\weight_T$ correspond to the dynamic programming equations of DTW
%for the computation of an optimal match between words,
%the matching cost for two symbols being the
%the time distance between them.
%\endex
%\end{example}

@@

%\subsection{Definition}
\noindent
\label{sec:SWVPA-def}
Let $\Delta$ be a countable alphabet
%finite (large) or infinite,
partitioned into three
subsets~$\Deltai$, $\Deltac$, $\Deltar$,
whose elements are respectively called
\emph{internal}, \emph{call} and \emph{return} symbols~\cite{AlurMadhusudan09nested}.
% names are  coined by application to functional program verification
% \begin{itemize}
% \item a set $\Deltai$ of \emph{internal symbols} denoted $a$,
% \item a set $\Deltac$ of \emph{call symbols} denoted $\call{a}$,
% \item a set $\Deltar$ of \emph{return symbols} denoted $\return{a}$.
% \end{itemize}
Let~$\< \Semiring, \oplus, \zero, \otimes, \one>$ be a commutative and complete semiring and let
$\bar\Phi = \< \Phii, \Phic, \Phir, \Phici, \Phicc, \Phicr>$
be a label theory over $\Semiring$
%In order to simplify notations, %and following the definition of Section~\ref{section:transducer},
%we shall write respectively
where $\Phii$, $\Phic$, $\Phir$ and~$\Phicx$ (with $\mathsf{x} \in \{ \mathsf{i}, \mathsf{c}, \mathsf{r}\}$)
stand respectively
for~$\Phi_\Deltai$, $\Phi_\Deltac$, $\Phi_\Deltar$ and~$\Phi_{\Deltac, \Delta_\mathsf{x}}$.
%
%Moreover, we extend this theory with a set $\Phii$
%of unary functions in $\Deltai \to \Semiring$,
%closed under $\oplus$ and $\otimes$.

@@

\begin{example} \label{ex:SWVPA}
We consider a \SWVPA over the alphabet of Example~\ref{ex:nested-word}
that expresses a weight related to the music notation,
or more precisely to its structural complexity.
Given a set of equivalent representations, we aim at choosing the simpler one.

%Let us assume a third component in the symbols of $\Deltac$
%\philippe{Je trouve l'exemple un peu confus. Je ne sais pas
%quel est "l'associate time interval" d'un symbole. Je ne sais
%pas à quoi fait référence "the following call transition". Je
%n'ose pas trop toucher mais je crois qu'il faut revoir.}
%which is the duration of the associated time interval.
%It is used to check the timestamps in the symbols.
%
For instance, the following call transition starts in state
$q_{{i}/{c}}$, meaning that the current interval has number~$i < c$
amongst $c$ sub-intervals of duration~$\ell$ (as indicated by the interval at the stack's top).
It reads a new time-division symbol~$\ccall{d}$:
\(
\weic\bigl(q_{{i}/{c}},
           \ccall{c}\ts{[\tau, \tau+\ell]}, q,
           \ccall{d}\ts{\iota}, q_{{i+1}/{c}},
           q_{{0}/{d}} \bigr) = \alpha_d
\).
The time interval $\iota$ attached to the~$\ccall{d}$ read 
must have a starting time $\tau + \frac{i\ell}{c}$,
where~$\tau$ and~$\ell$ are respectively
the starting time and duration of the interval attached to
the previous time-division symbol read $\ccall{c}$ (found on the stack).
%plus the total duration of intervals read so far, $i \ell$, and
%the new current interval duration is~$\frac{\ell}{d}$.
%
And it must have a duration $\frac{\ell}{d}$.

@@

%\subsection{Properties}
\medskip\noindent
Similarly to \VPA~\cite{AlurMadhusudan09nested}
and \SVPA~\cite{dAntonyAlur14SVPDA},
the class of \SWVPA is closed under the binary operators of the underlying semiring.
%
\begin{proposition}\label{prop:SWVPA-product}
Let $A_1$ and $A_2$ be two \SWVPA
over the same $\Delta$, commutative $\Semiring$ and effective $\bar\Phi$.
There exist two effectively constructible $\SWVPA$
$A_1 \oplus A_2$ and $A_1 \otimes A_2$,
such that for every $s \in \Delta^*$,
$(A_1 \oplus A_2)(s) = A_1(s) \oplus A_2(s)$ and
$(A_1 \otimes A_2)(s) = A_1(s) \otimes A_2(s)$.
\end{proposition}
%
\begin{proof}
We do a classical product construction, see Appendix~\ref{sec:product} for details.
%similar to the case of the Boolean semiring~\cite{dAntonyAlur14SVPDA}.
\end{proof}
\noindent
We present now a procedure for searching, for a~$\SWVPA$ $A$,
a word %$w \in \Delta^*$
of minimal weight for~$A$.
%as stated in the following proposition. %\wrt~$\leq_\oplus$,
%It requires the following property of label theories.
%
\begin{proposition}\label{th:best-search}
For a \SWVPA $A$
over $\Delta$,
$\Semiring$ commutative, bounded, total and complete, %semiring
and $\bar\Phi$ effective, %and $k$-convex, % label theory,
one can construct in PTIME a word $t \in \Delta^*$
such that $A(t)$ is minimal \wrt the natural ordering $\leq_\oplus$ for $\Semiring$.
\end{proposition}
%\florent{total?}
%
Let $A = \< Q, P, \init, \bar\wei, \final >$.
%
We propose a Dijkstra-like algorithm computing,
for every $q, q' \in Q$,
the minimum, \wrt $\leq_\oplus$, of the function
$\beta_{q, q'} : t \mapsto \weight_A(\config{q}{\bot}, t, \config{q'}{\bot})$.
Let us denote by $b_\bot(q, q')$ this minimum.
By definition of $\leq_\oplus$, and since $\Semiring$ is total,
it holds that:
%
\begin{equation}\label{eq:bbot}
  b_\bot(q, q') = \bigoplus_{t\in \Delta^*}
  \textstyle
  \weight_A\bigl(\config{q}{\bot}, t, \config{q'}{\bot}\bigr).
\end{equation}
The infinite sum in~\eqref{eq:bbot} is well defined since $\Semiring$ is complete.
%
Following~\eqref{eq:SWVPA-value}, and the associativity, commutativity
and distributivity for~$\otimes$ and~$\oplus$, the minimum of~$A(t)$ is:
\begin{equation}\label{eq:min}
{\displaystyle\bigoplus_{t\in \Delta^*}} A(t)
=
{\displaystyle\bigoplus_{t\in \Delta^*}}
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
\beta_{q, q'}(t)
\mathop{\otimes} \mathsf{out}(q')
=
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
b_\bot(q, q')
\mathop{\otimes} \mathsf{out}(q')
\end{equation}

>>>>

% comments in margin with \todo command
\usepackage[colorinlistoftodos]{todonotes} % option disable to hide todos
% command todo : options
% - noline: no line connecting the note with the place in the text
%   where the note occurs in the latex code
% - fancyline : curved arrow from note to inssertion point
% - inline: place a todonote inside the text instead of in the margin
% - author=name
\newcommand{\florent}[1]{\todo[noline,size=\tiny,color=yellow!40]{#1}}
%\renewcommand{\florent}[1]{} % uncomment for final
\newcommand{\lydia}[1]{\todo[noline,size=\tiny,color=red!30]{#1}}
%\renewcommand{\lydia}[1]{} % uncomment for final
\newcommand{\philippe}[1]{\todo[noline,size=\tiny,color=green!30]{#1}}
%\renewcommand{\philippe}[1]{} % uncomment for final
\newcommand{\reviews}[1]{\todo[noline,size=\tiny,color=blue!30]{#1}}
%\renewcommand{\reviews}[1]{} % uncomment for final

@@

\begin{comment}
%\begin{example}
%We can define with a \SWT the computation of a similarity measure
%between timed sequences
%similar to dynamic time warping (DTW).
%
%Let $\Semiring$  be the tropical (\emph{min-plus}) semiring of Figure~\ref{fig:semirings} and
%let $\Sigma = \Delta = \mathbb{R}_+$ be sets of timestamps.
%We consider a \SWT with one state $q$ and transitions
%$\wei_{11}(q, d, d', q) =
% \wei_{01}(q, d, d', q) =
% \wei_{10}(q, d, d', q) = |d' - d|$,
%for all $d, d' \in \mathbb{R}_+$.
%% needs reading input/output symbols even by epsilon-transitions.
%The recursive definition of $\weight_T$ correspond to the dynamic programming equations of DTW
%for the computation of an optimal match between words,
%the matching cost for two symbols being the
%the time distance between them.
%\endex
%\end{example}
\end{comment}

@@

%\subsection{Definition}
\noindent
\label{sec:SWVPA-def}
\reviews{3) pb countable}
Let $\Delta$ be a countable alphabet,
%finite (large) or infinite,
partitioned into three
pairwise disjoint subsets~$\Deltai$, $\Deltac$, $\Deltar$,
whose elements are respectively called
\emph{internal}, \emph{call} and \emph{return} symbols~\cite{AlurMadhusudan09nested}.
% names are  coined by application to functional program verification
% \begin{itemize}
% \item a set $\Deltai$ of \emph{internal symbols} denoted $a$,
% \item a set $\Deltac$ of \emph{call symbols} denoted $\call{a}$,
% \item a set $\Deltar$ of \emph{return symbols} denoted $\return{a}$.
% \end{itemize}
Let~$\< \Semiring, \oplus, \zero, \otimes, \one>$ be a commutative and complete semiring and let
$\bar\Phi = \< \Phii, \Phic, \Phir, \Phici, \Phicc, \Phicr>$
be a label theory over $\Semiring$
%In order to simplify notations, %and following the definition of Section~\ref{section:transducer},
%we shall write respectively
where $\Phii$, $\Phic$, $\Phir$ and~$\Phicx$ 
(where $\mathsf{x}$ is either $\mathsf{i}$, $\mathsf{c}$, or $\mathsf{r}$)
stand respectively
for~$\Phi_\Deltai$, $\Phi_\Deltac$, $\Phi_\Deltar$ and~$\Phi_{\Deltac, \Delta_\mathsf{x}}$.
%
%Moreover, we extend this theory with a set $\Phii$
%of unary functions in $\Deltai \to \Semiring$,
%closed under $\oplus$ and $\otimes$.

@@

\begin{example} \label{ex:SWVPA}
\reviews{revision ex. needed. see rev. 3) and app. $\weic$, $\weir$, $\weii$ wrong.}
We consider a \SWVPA over the alphabet of Example~\ref{ex:nested-word}
that expresses a weight related to the music notation,
or more precisely to its structural complexity.
Given a set of equivalent representations, we aim at choosing the simpler one.

%Let us assume a third component in the symbols of $\Deltac$
%\philippe{Je trouve l'exemple un peu confus. Je ne sais pas
%quel est "l'associate time interval" d'un symbole. Je ne sais
%pas à quoi fait référence "the following call transition". Je
%n'ose pas trop toucher mais je crois qu'il faut revoir.}
%which is the duration of the associated time interval.
%It is used to check the timestamps in the symbols.
%
For instance, the following call transition starts in state
$q_{{i}/{c}}$, meaning that the current interval has number~$i < c$
amongst $c$ sub-intervals of duration~$\ell$ (as indicated by the interval at the stack's top).
It reads a new time-division symbol~$\ccall{d}$:
\(
\weic\bigl(q_{{i}/{c}},
           \ccall{c}\ts{[\tau, \tau+\ell]}, q,
           \ccall{d}\ts{\iota}, q_{{i+1}/{c}},
           q_{{0}/{d}} \bigr) = \alpha_d
\).
The time interval $\iota$ attached to~$\ccall{d}$ 
must have a starting time $\tau + \frac{i\ell}{c}$,
where~$\tau$ and~$\ell$ are respectively
the starting time and duration of the interval attached to
the previous time-division symbol read $\ccall{c}$ (found on the stack).
%plus the total duration of intervals read so far, $i \ell$, and
%the new current interval duration is~$\frac{\ell}{d}$.
%
And it must have a duration $\frac{\ell}{d}$.

@@

%\subsection{Properties}
\medskip\noindent
Similarly to \VPA~\cite{AlurMadhusudan09nested}
and \SVPA~\cite{dAntonyAlur14SVPDA},
the class of \SWVPA is closed under the binary operators of the underlying semiring.
%
\begin{proposition}\label{prop:SWVPA-product}
Let $A_1$ and $A_2$ be two \SWVPA
over the same $\Delta$, commutative $\Semiring$ and effective $\bar\Phi$.
There exist two effectively constructible $\SWVPA$
$A_1 \oplus A_2$ and $A_1 \otimes A_2$,
such that for every $s \in \Delta^*$,
$(A_1 \oplus A_2)(s) = A_1(s) \oplus A_2(s)$ and
$(A_1 \otimes A_2)(s) = A_1(s) \otimes A_2(s)$.
\end{proposition}
%
\begin{proof}
We do a classical product construction, see Appendix~\ref{sec:product} for details.
%similar to the case of the Boolean semiring~\cite{dAntonyAlur14SVPDA}.
\end{proof}
\noindent
We present now a procedure for searching, for a~$\SWVPA$ $A$,
a word %$w \in \Delta^*$
of minimal weight for~$A$.
%as stated in the following proposition. %\wrt~$\leq_\oplus$,
%It requires the following property of label theories.
%
\begin{proposition}\label{th:best-search}
For a \SWVPA $A$
over $\Delta$,
$\Semiring$ commutative, bounded, total and complete, %semiring
and $\bar\Phi$ effective, %and $k$-convex, % label theory,
one can construct in PTIME a word $t \in \Delta^*$
such that $A(t)$ is minimal \wrt the natural ordering $\leq_\oplus$ for $\Semiring$.
\end{proposition}
%\florent{total?}
%
Let $A = \< Q, P, \init, \bar\wei, \final >$.
%
We propose a Dijkstra-like algorithm computing,
for every $q, q' \in Q$,
the minimum, \wrt $\leq_\oplus$, of the function
$\beta_{q, q'} : t \mapsto \weight_A(\config{q}{\bot}, t, \config{q'}{\bot})$.
Let us denote by $b_\bot(q, q')$ this minimum.
By definition of $\leq_\oplus$, and since $\Semiring$ is total,
\reviews{1) $\Semiring$ total $\to$ $\Semiring$ extremal}
it holds that:
%
\begin{equation}\label{eq:bbot}
  b_\bot(q, q') = \bigoplus_{t\in \Delta^*}
  \textstyle
  \weight_A\bigl(\config{q}{\bot}, t, \config{q'}{\bot}\bigr).
\end{equation}
The infinite sum in~\eqref{eq:bbot} is well defined since $\Semiring$ is complete.
%
Following~\eqref{eq:SWVPA-value}, and the associativity, commutativity
and distributivity for~$\otimes$ and~$\oplus$, the minimum of~$A(t)$ is:
\begin{equation}\label{eq:min}
{\displaystyle\bigoplus_{t\in \Delta^*}} A(t)
=
{\displaystyle\bigoplus_{t\in \Delta^*}}
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
\beta_{q, q'}(t)
\mathop{\otimes} \mathsf{out}(q')
=
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
b_\bot(q, q')
\mathop{\otimes} \mathsf{out}(q')
\end{equation}
